Optimizing Automated Finite Element Integration through Expression Rewriting and Code Specialization

### 2014-08-08 TOMS_Paper/assembly.tex
Start writing the perf eval section

We evaluate the effect of three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied. In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning). This way, we also highlight the advances achieved over our previous work.

@@

Each figure reported in the following is a grid of plots. Each grid has two axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis. The top-left plot will show values for $q=1, p=1$; the plot on its right will be $q=1, p=2$ and so on. The diagonal of the grid shows plots for which both basis and coefficient functions originate from the same function space.

A plot reports speed-ups over non-optimized FEniCS-Form-Compiler-generated code. There are three groups of bars, each group referring to a particular version of the code (\texttt{ffc, fix, auto}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on. Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.

\subsection{Performance of Forms}
In Figure~\ref{fig:helmholtz}, ...

@@

In Figure~\ref{fig:elasticity}, ...

@@

@@

\subsection{Details on the Autotuning Process}

% Mass DGEMM utile explicit method etc

@@

We have presented ...

>>>>

We evaluate the \textit{speed ups} achieved by three sets of optimizations over the original code; that is, the code generated by the FEniCS Form Compiler when no optimizations are applied. In particular, we analyze the impact of the FEniCS Form Compiler's built-in optimizations (henceforth \texttt{ffc}), the impact of COFFEE's transformations as presented in~\cite{Luporini} (referred to as \texttt{fix}, in the following), and the effect of Expression Rewriting and Code Specialization as described in this work (henceforth \texttt{auto}, to denote the use of autotuning as described in Section~\ref{sec:autotune}). The \texttt{auto} values do not include the autotuner cost, which is commented aside in Section~\ref{sec:auto-analysis}.

%This way, we also highlight the advances achieved over our previous work.

@@

Each figure reported in the following corresponds to one specific problem and must be read as a grid of plots. Each grid has two axes: $p$ varies along the horizontal axis, while $q$ varies along the vertical axis. The top-left plot will show values for $[q=1, p=1]$; the plot on its right will be $[q=1, p=2]$ and so on. The diagonal of the grid shows plots for which both basis and coefficient functions originate from the same function space. Therefore, a grid can be read in many different ways, which allows us to make structured considerations on the effect of the various optimizations.

A plot reports speed-ups over non-optimized FEniCS-Form-Compiler-generated code. There are three groups of bars, each group referring to a particular version of the code (\texttt{ffc, fix, auto}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

\subsection{Performance of Forms}
The four chosen forms allow us to perform an in-depth evaluation of different classes of optimizations for local assembly. We limit ourselves to analyzing the cost of computing element matrices, although all of the techniques presented in this paper are immediately extendible to the evaluation of local vectors. As anticipated, in the following we comment speed ups of \texttt{ffc}, \texttt{fix}, and \texttt{auto} over the non-optimized, FEniCS-Form-Compiler-generated code.

We first comment on results of general applicability. We note there is a trend of COFFEE's optimizations to become more and more effective as $q$, $p$, and $nf$ increase. This is because most of the transformations applied aim at optimizing for arithmetic intensity and SIMD vectorization, which obviously have a strong impact when arrays and iteration spaces are large. The corner cases of this phenomenon are indeed $[q=1, p=1]$ and $[q=4, p=4]$. We also observe how \texttt{auto}, in almost all scenarios, outperforms all of the other variants. In particular, it is not a surprise that \texttt{auto} is faster than \texttt{fix}, since \texttt{fix} is one of the autotuner's tested variants, as explained in Section~\ref{sec:autotune}. This proves the quality of the work presented in this paper, which shows significant advances over~\cite{Luporini}. The reasons for which \texttt{auto} exceeds both original code and \texttt{ffc} are discussed for each specific problem next. Also, details on the ``optimal'' code variant determined by autotuning are given in Section~\ref{sec:auto-analysis}.

@@

The results for the Helmholtz problem are provided in Figure~\ref{fig:helmholtz}. We observe that \texttt{ffc} slows the code down, especially for $q \geq 3$, as a consequence of using indirection arrays in the generated code that, as explained in Section~\ref{sec:zeros}, prevent, among the other compiler's optimizations, SIMD auto-vectorization. The \texttt{auto} version results in minimal performance improvements over \texttt{fix} when $nf=0$, unless $q=4$. This is due to the fact that if the loop over quadrature points is relatively small, then close-to-peak performance is obtainable through basic expression rewriting and code specialization; in this circumstance, generalized loop-invariant code motion and padding plus data alignment. The trend changes dramatically as $nf$ and $q$ increase: a more ample spectrum of transformations must be considered to find the optimal local assembly implementation. We will comment on this in details in the next section.

@@

Figure~\ref{fig:elasticity} illustrates results for the Elasticity problem. This form uses vector-valued spaces for the basis functions, so here transformations avoiding computation over zero-valued columns are of key importance. The \texttt{ffc} set of optimizations leads to notable improvements over the original code at $q=1$. The use of inderection arrays allows to phisically eliminate zero-valued columns at code generation time; as a consequence, different tabulated basis functions are merged into a single array. Therefore, despite the execution being purely scalar because of indirection arrays, the reduction in arithmetic intensity and register pressure imply improvement in performance. Nevertheless, \texttt{auto} remains in general the best choice, with gains over \texttt{ffc} that are wider as $p$ and $nf$ increase.

For $q \geq 2$, in \texttt{ffc} the lack of SIMD vectorization counterbalances the decrease in the number of floating point operations, leading to speed ups over the original code that only occasionally exceeds 1$\times$. On the other hand, the successful application of the zero-avoidance optimization while preserving code specialization plays a key role for \texttt{auto}, resulting in much higher performance code especially at $q=2$ and $q=3$.

It is worth noting that speed ups of \texttt{auto} over \texttt{fix} decrease at $q=4$, particularly for low values of $p$. As we will discuss in Section~\ref{sec:auto-analysis}, this is because at $q=4$ the vector-register tiling transformation (in combination with loop unroll-and-jam) leads to the highest performance. In principle, vector-register tiling can be used in combination to the zero-avoidance technique; however, due to mere technical limitations, this is currently not supported in COFFEE. Once solved, we expect much higher speed ups in the $q=4$ regime as well.

@@

\subsection{Details on the Autotuning Process}
\label{sec:auto-analysis}

% selected variants
% cost of autotuner
% cached autotuning

@@

We have presented ...
...TODO...: straightforward extendibility to right-hand side assembly
% Mass DGEMM utile explicit method etc
